---
title: 'Testing the ecological hypothesis: Longitudinal case-control study of early
  childhood caries and salivary microbiome assembly supplementary section'
author: "Freida Blostein"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r, echo = FALSE, include=FALSE, warning=FALSE, message=FALSE}
#prj directory
prjdir<-gsub('/Code', '', here::here())
data.out<-file.path(prjdir, 'Data')
code.dir<-file.path(prjdir, 'Code')
output.out<-file.path(prjdir, 'Output')
#original file folders - COHRA
og.data<-file.path(prjdir, 'OGData', 'COHRA')
meta.data<-file.path(og.data, 'Metadata')
dada.data<-file.path(og.data, 'DADA2')
wgs.path=file.path(prjdir, 'OGData', 'COHRA', 'WGS')

#results
main.out<-file.path(output.out, 'Main')
supp.out<-file.path(output.out, 'Supplement')
dir.create(main.out)
dir.create(supp.out)

#options
keepDouble=TRUE

#a billion libraries
#data manipulation
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(stringr)
library(stringi)
library(forcats)
library(readr)
library(tibble)
library(openxlsx)
library(compareGroups)
library(psych)
library(mclust)
library(DirichletMultinomial)
library(aricode)

#data visualization
library(ggplot2)
library(ggalluvial)
library(ggpubr)
library(ggtree)
library(ggjoy)
library(ggrepel)
library(tidygraph); 
library(ggraph);
library(igraph)
library(ggvenn)
library(cowplot)
library(patchwork)
library(grid)
library(gridExtra)
library(gtable)

#microbiome
library(phyloseq)
library(phangorn)
library(ape)
library(msa)
library(SQMtools)

#knitr options
knitr::opts_chunk$set(echo = FALSE, include=FALSE, warning=FALSE, message=FALSE)

```

```{r read in data}
#read in phyloseq object
load(file.path(data.out, 'phyloseq', 'phyasv_meta.Rdata'))
ps_long=microbiome::transform(ps_meta_nodups, 'compositional')%>%psmelt()%>%mutate(ScientificName=paste(Genus, ifelse(is.na(Species), OTU, Species), sep=' '))

#extract taxa info
taxa=as.data.frame(ps_meta_nodups@tax_table)%>%mutate(ASV=row.names(.), ScientificName=paste(Genus, ifelse(is.na(Species), ASV, Species), sep=' '))

#metavisit data
load(file.path(data.out, 'Metadata', 'MetaVisit.Rdata'))
if(keepDouble==T){MV=MetaVisit}
if(keepDouble==F){MV=MetaVisit%>%filter(count!='Double; drop')}
```

```{r load data supervised and unsupervised}
unsupervised_files<-c(list.files(file.path(data.out, 'Network'), full.names = T), list.files(file.path(data.out, 'CST'), full.names = T)[-1])

unsupervised_files%>%purrr::map(function(file){load(file, envir = .GlobalEnv)})

clr_network$module.data<-left_join(clr_network$module.data, MV)
hellinger_network$module.data<-left_join(hellinger_network$module.data, MV)

#pick most important cst 
main_cst=CST_k6
#merge with metavisit and rename NA as unclassified
CST1<-left_join(main_cst, MV %>% select(FoxCavID, UniquePersonID, UniqueID, BabySubjectID, Site, Case, IncidentVisit2, Visit, count))
levels(CST1$CST.8)<-c(levels(CST1$CST.8), 'Unclassified')
CST1[is.na(CST1$CST.8),'CST.8']<-"Unclassified"

#read in RF data
load(file.path(data.out, 'RF', 'rfResults_hellinger_iVisit_FALSE_duplicates_FALSE.Rdata'))

#wgs
load(file.path(data.out, 'MS', 'kegg_and_taxa_slim.Rdata'))
#load humann3 object
load(file.path(data.out, 'H3', 'Humann3.Rdata'))
```

```{r visit labels}
#visit labels for plotting
visit_labels2<-c('Birth', '2mos', 'First tooth \n (mean age=9 months)', '12mos', '24mos', '36mos', '48mos', '60mos')
names(visit_labels2)<-c('2', '3', '4', '5', '7', '8', '9', '10')
visit_labelsV2=visit_labels2
names(visit_labelsV2)<-paste0('V', c('3', '5', '7', '8', '9', '10'))
```

```{r name CSTs}
cst_recodes<-data.frame('CST.8'=as.factor(c(1, 2, 3, 4, 5, 6)), CSTlabels=factor(c('Neisseria ASV12 - V. dispar', 'H. parainfluenzae - Neisseria ASV9','Streptococcus ASV1 dominated w/ Gemella ASV2', 'Streptoccous ASV1 dominated w/ G. elegans', 'Gemella ASV2 - H. parainfluenzae - Neisseria ASV9', 'Streptococcus ASV8 - Neisseria ASV12'), levels=c('Streptococcus ASV1 dominated w/ Gemella ASV2', 'Streptoccous ASV1 dominated w/ G. elegans','Gemella ASV2 - H. parainfluenzae - Neisseria ASV9', 'Streptococcus ASV8 - Neisseria ASV12','H. parainfluenzae - Neisseria ASV9', 'Neisseria ASV12 - V. dispar')))
```

# Description of study sample

We used a nested case-control design with incident sampling to sample sequential saliva samples from 99 ECC cases and 90 incidence-visit matched controls from the Center for Oral Health Research in Appalachia 2 cohort (COHRA2). Children in COHRA2 are followed from birth until 5 years of age with regularly scheduled study visits in which they both provide oral samples and are examined for dental decay. Children therefore enter the cohort at birth, and all children are at-risk for developing ECC upon entry into the cohort. At each visit in which cases were diagnosed, we selected a similar number of ECC-free controls from the risk-set of all individuals still at risk of ECC, i.e. all those individuals who have not yet been diagnosed with ECC.  Incidence density sampling does not preclude the re-selection of a control as a case at later time points; controls can also be selected as controls more than once. (Supplemental Figure 1, 2). In this analysis, one control became a case at a later age, and one control was selected as a control in two different risk-sets. 

All available saliva samples from the cases and controls from the visit in which they were selected and all preceding visits were pulled for 16S rRNA amplicon sequencing. Selected individuals occasionally missed visits, did not have a saliva sample available, or had a saliva sample which failed 16S amplicon quality control (Supplemental Table 1). Additionally, we randomly selected a subcohort of 15 cases presenting with enamel lesions at or after the 36-month visit and 15 corresponding controls. Incident-visit plaque and saliva samples from these 30 individuals were submitted for whole genome shotgun metagenomic sequencing.  Additionally, we randomly selected a subcohort of 15 cases presenting with enamel lesions at or after the 36-month visit and 15 corresponding controls. Incident-visit plaque and saliva samples from these 30 individuals were submitted for whole genome shotgun metagenomic sequencing.  

```{r TABLE 1}
source(file.path(code.dir, 'Analysis', 'sampleLoss.R'))
write.xlsx(lostSamplesTab, file.path(supp.out, 'SupplementalTable1.xlsx'), overwrite=T)
```

```{r FIG 2, fig.width=13.2, fig.height=8, include=F}
cairo_pdf(file.path(supp.out, 'SupplementalFigure1.pdf'), family='Arial', width=13.2, height=8)
h1a
dev.off()
h1a
```

# 16S amplicon bioinformatics

## Sequencing

Samples were sequenced over 3 runs (each with 4 plates). Each plate included a positive mock control (ZymoBIOMICS Microbial Community DNA Standard (Zymo Research cat# D6306) and 2 negative controls (extraction control and water control). We processed sequenced reads. Sequencing was done on the Illumina MiSeq platform, using a MiSeq Reagent Kit V2 500 cycles (Illumina cat# MS102-2003), according to the manufacturers instructions with modifications found in the Schloss SOP.  Accuprime High Fidelity Taq (Life Technologies cat # 12346094 ) was used instead of Accuprime Pfx supermix. PCR products are visualized using an E-Gel 96 with SYBR Safe DNA Gel Stain, 2% (Life technologies cat# G7208-02).  Libraries are normalized using SequalPrep Normalization Plate Kit (Life technologies cat # A10510-01) following the manufactures protocol for sequential elution.  The concentration of the pooled samples is determined using Kapa Biosystems Library Quantification kit for Illumina platforms (KapaBiosystems KK4824).  The sizes of the amplicons in the library are determined using the Agilent Bioanalyzer High Sensitivity DNA analysis kit (cat# 5067-4626).  The final library consists of equal molar amounts from each of the plates, normalized to the pooled plate at the lowest concentration. 	Library Preparation for Sequencing and Sequencing Libraries are prepared according to Illumina’s protocol for Preparing Libraries for Sequencing on the MiSeq (part# 15039740 Rev. D) PhiX (15%) was added in 16s amplicon sequencing to create diversity.   Sequencing reagents are prepared according to the Schloss SOP, custom read 1, read 2 and index primers are added to the reagent cartridge.  FASTQ files are generated for paired end reads. DNA was quantified using PicoGreen.																													
## DADA2 processing

```{r dada2 and preprocess info}
track_reads=readRDS(file.path(dada.data, 'track_postchimera_HOMD.rds'))
readlimit=1000
load(file.path(output.out, 'Preprocess', paste0('preprocessParamsStats_', readlimit, '.Rdata')))
load(file.path(output.out, 'Preprocess', paste0('filtersamples_', readlimit, '.Rdata')))
```

Reads were processed to amplicon sequence variants (ASVs) using DADA2 (version 1.14.1) [Callahan] and the Human Oral Microbiome Database (HOMD) version 15.2 (Supplementary text) [Escapa 2020]. After examining quality plots for forward and reverse reads, reads were trimmed at the 240th and 200th nucleotide position respectively, or at the first instance of a quality score <= to 11. The following parameters were used for filtering reads: a maximum of 2 expected errors for both forward and reverse reads and no ambiguous bases. 

Errors were learned (independently for each run) using 1 million bases.  The forward and reverse reads were denoised and then forward and reverse reads were merged together. 

At this point, the runs were merged together, and chimeras were removed using DADA2s removeBimeraDenovo commmand and the consensus method. The mean read number, prior to any sample or taxa filtering and including both positive and negative controls, was `r round(mean(track_reads))`, sd=`r round(sd(track_reads))`. A total of m=`r taxa_count$all` ASVs were identified.

Subsequently, reads were assigned taxonomic names to the species level using the assignTaxonomy and the addSpecies commands of DADA2. A modified version of the Human Oral Microbiome Database (HOMD) version 15.2 (modified by adding the exact sequences of the positive control samples) was used. 

## Sample and taxa filtering

To identify potential contaminant taxa we used the R package decontam, identifying contaminants using both the frequency-based method (PicoGreen quantification used as measure of concentration of amplified DNA), and the prevalence based method; thresholds set at `r paste(decontam.thresh, collapse=' and ')` respectively. Contaminant identification was performed independently within each batch. Of the original `r taxa_count$all` ASVs, `r taxa_count$all-taxa_count$decontam` ASVs were identified as contaminants and removed.

Next, we filtered out samples that had less than `r readlimit` reads (Supplemental Figure 4; red line at read limit=`r readlimit`). Of a total original `r sample_count$all['True Sample']` samples, `r sample_count$lessthan1000['True Sample']` samples had < `r readlimit`, reads, leaving `r sample_count$morethan1000['True Sample']` samples. Three of the true samples which passed quality control were missing all metadata and were excluded, leaving a total of `r `r sample_count$truesamples['True Sample']` in the analysis. After restricting to just the true samples (not controls) with > `r readlimit`, `r taxa_count$truesamples` ASVs remained. Calculation of diversity metrics (Shannon index, Chao1) was performed on these ASVs. 

```{r FIG 3, include=F}
cairo_pdf(file.path(supp.out, 'SupplementalFigure3.pdf'), family='Arial', width=13.2, height=8)
readcountsbyType_preDecontam
dev.off()
readcountsbyType_preDecontam
```

Of the original `r sample_count$all['Mock']` mock communities, `r sample_count$lessthan1000['Mock']` failed to amplify had less than `r readlimit `reads; the other mock communities in the same run on other plates amplified and were as expected (Supplemental Figure XX). None of the extraction controls had more than `r readlimit` reads, and one water control did, this water control had <3000 reads of highly common taxa in the analytic subset and likely reflects some cross-contamination (Supplemental Figure XX). 

```{r FIG 4, include=F}
cairo_pdf(file.path(supp.out, 'SupplementalFigure4.pdf'), family='Arial', width=13.2, height=8)
ps_mock_bar
dev.off()
ps_mock_bar
```


Finally, for the unsupervised and supervised analysis, we further restricted the ASV set to ASVs. We filtered out ASVs that were both 1) present in less than `r prev*100`% of the samples and 2) represented less than `r abund.filter*100`% of the relative abundance in samples in which they were present. This left a total of `r taxa_count$taxafilter` ASVs. 


# Supervised random forest

## Additional investigation of *Streptococcus* amplicons identity 
```{r randomForest}
imp<-lapply(rfFits_hellinger[c(1, 3)], function(x) as.data.frame(x$finalModel$importance)%>%mutate(ASV=row.names(x$finalModel$importance)))%>%bind_rows(.id="column_label")%>%mutate(Visit=as.integer(gsub('V', '', gsub('All taxa', '', column_label))), Month=ifelse(Visit=='5', '12 month', '24 month'))
#info on >mean in case | control
greater_mean=ps_long %>% filter(Visit %in% c(5, 7))%>%group_by(OTU, Visit, Case, ScientificName)%>%summarise(mean=mean(Abundance))%>%pivot_wider(id_cols=c(OTU, Visit, ScientificName), names_from=Case, names_prefix='Abundance_', values_from=mean)%>%mutate(GreaterMean=ifelse(Abundance_Case>Abundance_Control, 'Case', 'Control'), ASV=OTU)
#join 
imp=left_join(imp, greater_mean)
imp_top=rbind(imp%>%group_by(Visit)%>%slice_max(n=10, order_by=MeanDecreaseGini), imp%>%filter(str_detect(ScientificName, 'mutans')))
```


```{r streptococcus blast}
mystreps=ps_meta_nodups@refseq[c('ASV8', 'ASV14', 'ASV82')]
Biostrings::writeXStringSet(mystreps, file.path(data.out, 'Reference', 'mystreps.fasta'))
rm(mystreps)
#run blast in web browser or terminal on mystreps.fasta

#load in description tables
files=list.files(path=file.path(data.out, 'Reference', 'BLAST'), pattern="*.csv", full.names=T)
strep_blast=sapply(files, read_csv, simplify=FALSE)%>%bind_rows(.id="id")
strep_blast=strep_blast%>%mutate(asv=gsub('.*/(asv[0-9]{1,2}).*.csv', '\\1', strep_blast$id))%>%select(asv, everything())%>%select(-id)
```

```{r TABLE 2 3 4}
write.xlsx(strep_blast %>% split(.$asv),
           file.path(supp.out, 'SupplementaryTables2-4.xlsx'), overwrite=T)
```

```{r streptococcus correlation plots}
h3_vs_16S=left_join(species_info$species_long, 
                    ps_long%>%
                      select(-Species, -SampleType, -CaseStatus), 
                    by=c('COHRAID', 'Visit'))%>%
            mutate(case_ever_numeric=ifelse(CaseEver=='Case', 1, 0))

ms_species=left_join(as.data.frame(t(saliva_species_percent))%>%rownames_to_column(var='SampleName')%>%pivot_longer(cols=-SampleName, names_to='Species', values_to='WGSAbundance'), species_info$sample_table)

ms_vs_16S=left_join(ms_species, ps_long%>%select(-Species, -SampleType, -CaseStatus), by=c('COHRAID', 'Visit'))%>%mutate(case_ever_numeric=ifelse(CaseEver=='Case', 1, 0))

cor_partial<-function(df){
  #partial spearman correlation to control for case status and visit
  #Resources: 
  #https://towardsdatascience.com/keeping-an-eye-on-confounds-a-walk-through-for-calculating-a-partia  l-correlation-matrix-2ac6b831c5b6
  #http://sherrytowers.com/2013/09/23/correlations-and-partial-correlations/
  #code from psych package, namely: http://personality-project.org/r/html/partial.r.html
  #https://www.rdocumentation.org/packages/psych/versions/2.0.12/topics/corr.test
  #n for sample size in corr.test is n-s where s is the number of variables partialed out, here I      used n-2 (CaseEver & Visit)= 30-2=28
  rho<-corr.p(partial.r(data=df, x=cs(WGSAbundance, Abundance), y=cs(case_ever_numeric, Visit), 
                        method='spearman'), n=28)
  return(data.frame('rho'=rho$r[2], 'pval'=rho$p[2]))}

calculateWGS16Scor=function(df, genus, asv_list=NA){
  if(!is.na(asv_list)){new_df=df %>% filter(str_detect(Species, genus) & OTU %in% asv_list)}
  if(is.na(asv_list)){new_df=df %>% filter(str_detect(Species, genus) & str_detect(Genus, genus))}
  
  grouped_df=group_by(new_df, SampleType, Species, ScientificName)%>%nest()%>%
    mutate(cor=purrr::map(data, cor_partial))%>%unnest(cor)%>%
    mutate(rho_ifsig=ifelse(pval<0.05, round(rho, 2), NA))
  new_plot=ggplot(grouped_df, aes(x=ScientificName, y=Species, fill=rho, label=rho_ifsig))+
    geom_tile()+geom_text()+facet_wrap(~SampleType)+xlab('16S amplicon')+ylab('WGS metagenomics')+
    theme_classic()+
    scale_fill_gradient2(name='Partial Spearman rho', mid="#FBFEF9",low="#0C6291",high="#A63446", 
                         limits=c(-1, 1)) +
    theme(axis.text=element_text(face='italic'))+theme(axis.text.x=element_text(angle=45, hjust=1))
  return(new_plot)
}

#filter to only saliva for strep plot in H3 & remove streps that are only in plaque
strep_nosaliva=h3_vs_16S%>%filter(SampleType=='Saliva')%>%group_by(Species)%>%summarise(sum=sum(WGSAbundance))%>%filter(sum==0)%>%pull(Species)
saliva_only=h3_vs_16S%>%filter(SampleType=='Saliva' & !(Species %in% strep_nosaliva))

#put plots together
strep_MS=calculateWGS16Scor(ms_vs_16S, 'Streptococcus', c('ASV1', 'ASV10', "ASV8", 'ASV14', 'ASV82', 'ASV26', 'ASV76', 'ASV196'))
strep_H3=calculateWGS16Scor(saliva_only, 'Streptococcus', c('ASV1', 'ASV10', "ASV8", 'ASV14', 'ASV82', 'ASV26', 'ASV76', 'ASV196'))
strep_cors=strep_H3/strep_MS+plot_annotation(tag_levels='A')
#save supplemental plot
ggsave(strep_cors, file=file.path(supp.out, 'SupplementalFigure_strepcors.eps'), width=8, height=10)
#remove large objects to preserve memory
rm(saliva_only, ms_species, h3_species)
```

```{r FIG 5, fig.width=8, fig.height=10, include=F}
ggsave(strep_cors, file=file.path(supp.out, 'SupplementalFigure5.eps'), width=8, height=10)
strep_cors
```

```{r diversity and smutans statistics}
div=readRDS(file.path(data.out, 'phyloseq', 'diversityandsmut.Rds'))
```

```{r TABLE 5}
network_data=hellinger_network$module.data%>%ungroup()%>%unique()%>%
  select(-module_label3, -module, -CentralTaxa1, -CentralTaxa2,
         -module_label, -TopTaxa1,-TopTaxa2,-moduleColor)%>%
  pivot_wider(names_from=module_label2, values_from=modAbund)

table1=left_join(MV, left_join(network_data, 
                               left_join(main_cst, cst_recodes)))%>%
                  left_join(div%>%
              dplyr::select(FoxCavID, Chao1, Shannon, smutAbundance,
                            smutPrevalence))

table1.noincident=table1 %>%filter(Visit < IncidentVisit2)
table2.incident=table1 %>% filter(Visit==IncidentVisit2)

res2<-compareGroups(Case~Shannon+Chao1+`Streptococcus ASV1 & Neisseria ASV12 network`+`Haemophilus parainfluenzae & Neisseria ASV9 network`+`Veillonella ASV5 & Streptococcus ASV8 network`+`Veillonella sp._HMT_780 & Granulicatella elegans network`+`Gemella ASV2 & Leptotrichia shahii network`+CSTlabels+smutPrevalence+smutAbundance,data=table2.incident%>%filter(Visit%in% c( 5, 7, 8)), include.miss=F)
strata.res2=strataTable(createTable(res2), 'Visit')
visits=c(5, 7, 8, 9, 10)
res_list=list()
for(i in 1:length(visits)){
  data=table2.incident%>% filter(Visit==visits[i])
  res=createTable(compareGroups(Case~Shannon+Chao1+`Streptococcus ASV1 & Neisseria ASV12 network`+`Haemophilus parainfluenzae & Neisseria ASV9 network` +
                `Veillonella ASV5 & Streptococcus ASV8 network`+`Veillonella sp._HMT_780 & Granulicatella elegans network`+
                  `Gemella ASV2 & Leptotrichia shahii network`+CSTlabels+smutPrevalence+smutAbundance,data=data, include.miss=F))
  res_list[i]=res
  #export2xls(res, file=file.path(supp.out, paste0('SupplementalTable2_v', visits[i], '.xlsx')))
}
names(res_list)=paste0(visit_labels2[names(visit_labels2)%in% visits])
write.xlsx(res_list, file=file.path(supp.out, 'SupplementalTable5.xlsx'), overwrite=T, rowNames=T)

```


```{r strep asv identity}
asv8_identity=paste(strep_blast %>%filter(asv%in% c('asv8') & `Per. ident`==100)%>%pull(`Scientific Name`)%>%unique(), collapse=' and ')
asv14_identity=paste(strep_blast %>%filter(asv%in% c('asv14') & `Per. ident`==100)%>%pull(`Scientific Name`)%>%unique(), collapse=' and ')
asv82_identity=paste(strep_blast %>%filter(asv%in% c('asv82') & `Per. ident`==100)%>%pull(`Scientific Name`)%>%unique(), collapse=' and ')
```

Species-level identification of *Streptococcus* with the V4 amplicon of the 16S gene is not perfect. Two amplicons identified as *Streptococcus* were important features in the taxa-wide random-forest, but did not identify to the species level. The identity of the amplicon identified as *Streptococcus mutans* is central to our test of the ecological hypothesis. We performed several analyses to improve our confidence in the taxonomic classification of these important *Streptococcus* amplicons. 

*Streptococcus ASV82*, which identified as *S. mutans*, had 100% identity with *`r asv82_identity`* in a BLAST search. 6 of the top 10 top hits were *S mutans*. None of the other 4 top 10 hits were human oral *Streptococcus* species (Supplemental Table 2). 

Neither *Streptoccocus ASV8* nor *Streptococcus ASV14* had a high identity with any *S. mutans* 16S sequences when performing a BLAST search. *Streptococcus ASV8* had 100% identity with *`r asv8_identity`*, *Streptococcus ASV14* with *`r asv14_identity`*. Neither *Streptococcus ASV8* or *ASV14* had an *S. mutans* BLAST hit in the top 100 best blast hits (Supplemental Tables 3 & 4). 

Additionally, we exploited our subcohort with metagenomically sequenced saliva samples to interrogate the identity of the *Streptococcus* amplicons. Abundance of the 16S amplicon *Streptococcus ASV82* (*S. mutans*) was strongly and significantly correlated with abundances of *S. mutans* in WGS samples when using either Humann3 profile abundances or abundances estimated from an assembly based pipeline, while amplicon abundances of *Streptococcus ASV8* correlated strongly and significantly with *S. salivarius* and *Streptococcus ASV14* with *S. australis* and *S. infantis* (Supplemental Figure 5, Pearson r values significant at *P* value<0.05 shown as text in the heatmap). 

# Unsupervised clustering techniques

## Community state typing: choice of k community state types

Using Dirichlet multinomial community state typing, we identified 6 community state types (also called enterotypes). We chose k=6 community state types as this choice of k produced the best model fit as measured by the LaPlace metric (Supplemental Figure XX). Each sample is assigned to a single community state type based on a posterior probability; any sample which did not assign to a community state type at a posterior probability >80% was labeled unclassified. Community state types were named by the taxa defining their separation. 

Supplemental Figure 7 shows the relationship between Shannon diversity and community state type. Community state type is significantly associated with Shannon diversity. As children age, the dominant community state types progress from those that are *Streptococcus* dominated to community state types that are higher diversity. The progression of community state types in both cases and controls follows this pattern. Within ages, cases and controls are more likely to belong to different community state types (Figure 2A; Supplemental Figure 7; community state types prevalent in cases in orange shades, community state types prevalent in controls in blue shades).


```{r FIG 6, include=F}
load(file.path(og.data, 'DMM_fit_asvtrim.rda'))
### The return value can be queried for measures of fit (Laplace, AIC, BIC); these are plotted for different k
fit<-fit_asvtrimmed
lplc <- sapply(fit, laplace)
plot(lplc, type="b", xlab="Number of Dirichlet Components", ylab="Model Fit", main='Laplace model fit at k=1:10 for community state typing')
cairo_pdf(file.path(supp.out, 'SupplementalFigure6.pdf'), family='Arial', width=13.2, height=8)
plot(lplc, type="b", xlab="Number of Dirichlet Components", ylab="Model Fit", main='Laplace model fit at k=1:10 for community state typing')
dev.off()
```

```{r FIG 7, include=F}
fig7=table1%>%ggplot(aes(x=CSTlabels, y=Shannon, color=CSTlabels))+geom_boxplot()+geom_jitter(alpha=0.2)+facet_wrap(~Visit,nrow=1, labeller=labeller(Visit=visit_labels2))+stat_compare_means(label='..p.signif..')+scale_color_manual(name="Community state type", values=c("thistle", "lightcoral", "deepskyblue",   "darkorange", "royalblue2",  "darkorange3",  "royalblue4", "darkorange4", 'grey', 'black'), guide=guide_legend(nrow=2))+theme_classic()+theme(axis.text.x = element_blank())+theme(legend.text=element_text(face='italic'))+theme(legend.position='bottom')+xlab('')#+stat_compare_means(label='p.signif', label.y=4, comparisons=list(c("Gemella ASV2 - H. parainfluenzae - Neisseria ASV9", "Streptococcus ASV8 - Neisseria ASV12"), c("H. parainfluenzae - Neisseria ASV9", "Neisseria ASV12 - V. dispar")))
fig7
cairo_pdf(file.path(supp.out, 'SupplementalFigure7.pdf'), family='Arial', width=13.2, height=8)
fig7
dev.off()
```

## Weighted co-occurence network analysis 

```{r unsupervised guilds networks - fig 2b}
#Pick top genera to color
topGenus<-c('Streptococcus', 'Neisseria', 'Veillonella', 'Fusobacterium', 'Haemophilus', 'Prevotella', 'Gemella', 'Actinomyces')
genus_colors=c(scales::hue_pal()(8), 'black')
names(genus_colors)=c(topGenus, 'Other')
#Pick top taxa to label 
topTax<- c('Fusobacterium periodonticum', 'Haemophilus parainfluenzae', 'Prevotella nanceinsis', 'Neisseria ASV9', 'Veillonella ASV5', 'Streptococcus ASV8', 'Prevotella histcola', 'Prevotella salivae', 'Prevotella melaninogenica', 'Streptococcus mutans', 'Fusobacterium nucleatum_subsp._animalis', imp_top%>%pull(ScientificName), hellinger_network$intraStats%>%group_by(module)%>%slice_max(n=2, order_by=kWithin)%>%pull(ScientificName))
#load extra fonts
extrafont::loadfonts()
#abundance info
my_df=ps_long%>% group_by(OTU,CaseEver)%>%summarise(m=mean(Abundance))%>%pivot_wider(names_from = CaseEver, values_from=m)%>%mutate(as=ifelse(Case>Control, 'Higher mean abundance case', 'Higher mean abundance controls'), m=ifelse(Case>Control, Case, Control))%>%mutate('name'=OTU)
#source tidygraph object creation from network object creation
nw<-hellinger_network
source(file.path(code.dir, 'Analysis', 'createTidyGraph.R'))


network_n=hellinger_network$module.data%>%filter(moduleColor %in% c('brown', 'yellow') & Visit<=IncidentVisit2)%>%group_by(moduleColor, Visit, Case)%>%summarise(n=n())%>%pivot_wider(names_from=Case, values_from=n)%>%mutate(modAbund=1, label=paste0('Case n=', Case, '\n Control n=', Control), Case='case')%>%filter(Visit %in% c(3, 5, 7, 8, 9, 10))%>%mutate(Visit=as.factor(Visit))

#plot change over time 
network_overtime_others=ggplot(hellinger_network$module.data%>%filter(Visit %in% c(3, 5, 7, 8, 9, 10) & !(moduleColor %in% c("brown", 'yellow')) & Visit<=IncidentVisit2), aes(x=as.factor(Visit), y=modAbund, color=as.factor(Case)))+stat_summary(geom='point', fun.y=mean)+facet_wrap(module_label3~., ncol=2, scales='free_y')+
    stat_summary(fun.data = mean_cl_boot, geom='errorbar', width=0.5)+#stat_compare_means(aes(label = ..p.signif..), size=5, hide.ns = TRUE, label.y=0.50)+
    geom_line(aes(group=UniquePersonID), alpha=0.1)+geom_smooth(aes(group=as.factor(Case), color=as.factor(Case)), se=F)+scale_color_manual(values=c("darkorange3", "royalblue3"), labels=c("ECC case", "Control"), name="")+theme_classic()+
    scale_x_discrete(name="Age (months)", labels=visit_labels2, expand=c(.1, .1))+ylab("Summed module relative abundance")+
    theme(text=element_text(size=14))+theme(axis.text.x=element_text(angle = 45, vjust = 0.5, hjust=.5))+
    #scale_y_continuous(breaks=c(0, .5, 1), minor_breaks = c(.25, .75))+
    theme(legend.position='bottom')+theme(strip.text = element_text(face='italic'))
```

```{r FIG 8, include=F}
network_overtime_others
ggsave(filename=file.path(supp.out, 'SupplementalFigure8.eps'), network_overtime_others, units=c('in'), width=10, height=10, device=cairo_ps)
```

```{r FIG 9, include=F}
entire_network<-gg_whole%>%ggraph('nicely')+geom_edge_link(aes(width=ifelse(weight>0.1, weight, NA)), alpha=1, color='grey')+geom_node_point(aes(color=factor(module_label3), size=m, shape=factor(as)))+geom_node_text(aes(label=ScientificNameDisplay), fontface='italic', family='Times', repel=TRUE, size=3)+guides(size=FALSE, edge_color=FALSE, color=guide_legend(nrow=2,byrow=TRUE), edge_width=FALSE)+scale_shape_manual('', values=c('triangle', 'square'))+scale_color_discrete(name='')+theme_graph(base_family='serif')+theme(legend.position = 'bottom')+scale_edge_width(range = c(1, 8))+scale_size(range=c(2, 8))+theme(strip.text.x = element_blank())+theme(text=element_text(family='serif', face='italic'))

entire_network

ggsave(filename=file.path(supp.out, 'SupplementalFigure9.eps'), entire_network, units=c('in'), width=10, height=10, device=cairo_ps)
```

```{r TABLE6}
rm(entire_network)
module_membership<-left_join(hellinger_network$intraStats, hellinger_network$networkNames)%>%select(ASV, ScientificName, module_label2, module_label3,, module, kTotal, kWithin, kOut, kDiff)%>%group_by(module)%>%split(f=as.factor(.$module))
write.xlsx(module_membership, file=file.path(supp.out, 'SupplementalTable6.xlsx'), overwrite = T)
```


Using weighted co-occurrence network analysis, we identified 5 network modules of co-occurring ASVs. These networks were named after the top two most abundant taxa in the network, and the most central taxa in the network. To summarize these network modules, we summed the relative abundances of each taxa belonging to the network module compared the trajectories of these abundances over time between cases and controls. Change in the summed relative abundance of the 3 network modules not displayed in Figure 2B are shown in Supplemental Figure XX; the structure of the overall network is shown in Supplemental Figure XX. 

## Overlap between top ten separation taxa from community state typing and top two most abundant taxa from each network modules

```{r unsupervised guilds vendiagram}
tax_contribution<-left_join(read.csv(file.path(output.out, 'CST', 'k6', 'k_6taxcontribution.csv'), col.names = c('ASV', 'Mean', paste0('CST', seq(1, 6)), 'diff', 'cdiff')), taxa)

#get the names of the top separation taxa (Cst)
#& top abundant network and hub taxa
top_seps=tax_contribution%>%head(n=10)%>%pull(ScientificName)
top_nets=c(hellinger_network$networkNames$TopTaxa1, hellinger_network$networkNames$TopTaxa2)
top_hubs=c(hellinger_network$networkNames$CentralTaxa1, hellinger_network$networkNames$CentralTaxa2)

#Get intersections for venndiagram 
middle=top_seps[top_seps %in% top_nets & top_seps %in% top_hubs]
seps_nets=top_seps[top_seps %in% top_nets]
seps_hubs=top_seps[top_seps %in% top_hubs]
nets_hubs=top_hubs[top_hubs %in% top_nets]

#make venn diagram 
width_line<-1
#could do venn diagram comparing hubs to top abundant taxa +seps
#ggvenn(list('Top 10 CST \n separation taxa'=top_seps, 'Top 2 most abundant \n taxa from each WCN'=top_nets, 'Top 2 most central \n taxa from each WCN'=top_hubs))
#but for this paper, specifically, think the comparison between 
#top most abundant network hubs and top separation taxa is more important 
venn_comp=ggvenn(list('Top 10 CST \n separation taxa'=top_seps, 'Top 2 most abundant \n taxa from each WCN'=top_nets), show_elements = T, text_size=2, label_sep='\n')
```

```{r FIG 10, include=F}
venn_comp

ggsave(filename=file.path(supp.out, 'SupplementalFigure10.eps'), venn_comp, units=c('in'), width=8, height=8)
```

## Robustness analysis

For the Dirichlet multinomial community state typing, we varied the choice of k community state types. When decreasing k to 5, the case-associated 12-month community state type was combined with one of the 2-month community state types (Supplemental Figure XX). When decreasing k to 4, the control-associated community state type was also collapsed. However, case- and control-associated community state types apparent at and after the 24-month visit were still apparent. The clustering of the samples through community state types was overall fairly robust, according to the rand index, the adjusted rand index and the adjusted mutual information index (Supplemental table XX).


```{r cst robust}
CST_compare<-left_join(left_join(left_join(CST_k6%>%mutate_at(vars(CST.8), function(x){paste0('k6_', x)})%>% rename_at(vars(CST_nocut:maxcomp.9), function(x){paste0("k6.", x)}),  CST_k5 %>%mutate_at(vars(CST.8), function(x){paste0('k5_', x)})%>% rename_at(vars(CST_nocut:maxcomp.9), function(x){paste0("k5.", x)})), CST_k4%>%mutate_at(vars(CST.8), function(x){paste0('k4_', x)}) %>% rename_at(vars(CST_nocut:maxcomp.9), function(x){paste0("k4.", x)})), MV %>% select(FoxCavID, UniquePersonID, UniqueID, BabySubjectID, Site, Case, IncidentVisit2, Visit))
compare=c('k4.CST.8', 'k5.CST.8', 'k6.CST.8')
comparedf=data.frame()
for(i in 1:(length(compare)-1)){
  for(j in 2:(length(compare))){
    if(i!=j){
    newrow=clustComp(CST_compare[, compare[i]], CST_compare[, compare[j]])%>%bind_rows()%>%
      mutate(comparison=paste0(compare[i], ' vs ', compare[j]), 
             cohort='COHRA2')
    comparedf=rbind(comparedf, newrow)
    }
  }
}

CST_compare_al<-CST_compare %>% group_by(Case, Visit, k6.CST.8, k5.CST.8, k4.CST.8) %>% dplyr::summarise(count=n())%>%mutate_at(vars(matches('CST.8')), ~ifelse(str_detect(., 'NA'), 'NA', .))

CST_compare_al<-to_lodes_form(CST_compare_al, axes=3:5, id='Cohort')

CST_compare_al$stratum=factor(CST_compare_al$stratum, levels=c('k6_3', 'k6_4', 'k6_6', 'k6_5', 'k6_1', 'k6_2', 'k5_4', 'k5_2', 'k5_5','k5_3', 'k5_1', 'k4_2', 'k4_1', 'k4_3', 'k4_4', 'NA'))

visit_labels=c('2 mo', '12 mo', '24 mo', '36 mo')
names(visit_labels)=c(3, 5, 7, 8)

cst_sens<-ggplot(CST_compare_al%>%filter(Visit %in% c(3, 5, 7, 8)),
       aes(x = x, stratum = stratum, alluvium = Cohort,
           y = count, fill=stratum)) +
    stat_stratum(width=.5)+geom_flow(alpha=.7)+geom_stratum(width=.5)+facet_grid(Case~Visit, labeller=labeller(Visit=visit_labels))+
    scale_fill_manual(name="", values=c("thistle", "lightcoral", "darkorange", "deepskyblue", "darkorange3","royalblue2", 'thistle', 'lightcoral', 'deepskyblue', 'darkorange3', 'royalblue2', 'thistle', 'lightcoral', 'darkorange3', 'royalblue2', 'grey'), labels=c(paste(rep('k=6: CST', 6), seq(1, 6)), paste(rep('k=5: CST', 5), seq(1, 5)), paste(rep('k=4: CST', 4), seq(1, 4)), 'Unassigned at P>0.8'))+scale_y_continuous('Number of samples assigned to CSTs')+scale_x_discrete(name='', labels=c('6 CSTs', '5 CSTs', '4 CSTs'))
```

```{r FIG 11, include=F}
cst_sens
ggsave(filename=file.path(supp.out, 'SupplementalFigure11.eps'), cst_sens, units=c('in'), width=8, height=8)
```

For the weighted co-occurrence network clusters, we varied the choice of normalization transform, comparing the Hellinger transform to the center-log ratio transform. When using the center-log ratio transform instead of the Hellinger transform, six network modules were identified instead of five. The network module with the most member taxa was split in two (Supplemental Figure XX), and some rearrangement of taxa membership existed between the other network modules, thus, clustering of the taxa across different choices of transform were better than random but not high. However, the ECC case status associated clusters (control enriched *Neisseria* ASV9-*H. parainfluenzae* network; case-enriched *Veillonella* ASV5-*Streptococcus* ASV8 network) were more stable and relatively robust (Supplemental Figure XX; Supplemental Table XX).

```{r network robustness}
#a dataframe of the taxa in each network cluster across the two different transformations
compareModules=data.frame(taxa, 'clr'=clr_network$taxaColors, 'hellinger'=hellinger_network$taxaColors)
#rename the modules
hellingerColors=setNames(hellinger_network$networkNames$module_label3, hellinger_network$networkNames$module)
clrColors=setNames(clr_network$networkNames$module_label3, clr_network$networkNames$module)

compareModules$clr<-dplyr::recode(compareModules$clr, !!!clrColors)
compareModules$hellinger<-recode(compareModules$hellinger, !!!hellingerColors)
#convert to lode form for alluvial plot
compMod<-compareModules %>% group_by(hellinger, clr)%>%dplyr::summarise(freq=n())
compMod_lodes<-to_lodes_form(compMod, axes=1:2)

#reorder strata to make comparison easier
compMod_lodes$stratum<-factor(compMod_lodes$stratum, levels=c(
  "Haemophilus parainfluenzae & Neisseria ASV9 network\n (central taxa: Fusobacterium periodonticum)", 
  "Veillonella sp._HMT_780 & Granulicatella elegans network\n (central taxa: Alloprevotella sp._HMT_473)", 
  "Veillonella ASV5 & Streptococcus ASV8 network\n (central taxa: Lachnoanaerobaculum orale)", 
  "Prevotella histicola & Rothia ASV49 network\n (central taxa: Megasphaera micronuciformis)",
  "Gemella ASV2 & Leptotrichia shahii network\n (central taxa: Prevotella sp._HMT_317)",
  "Neisseria ASV12 & Veillonella dispar network\n (central taxa: Actinomyces ASV41)", 
  "Bergeyella sp._HMT_206 & Rothia ASV34 network\n (central taxa: Capnocytophaga ASV119)",
  "Streptococcus ASV1 & Neisseria ASV12 network\n (central taxa: Actinomyces ASV41)",
  "Streptococcus ASV1 & Gemella ASV2 network\n (central taxa: Prevotella ASV479)"
))

#plot taxa change
network_sensitivity<-ggplot(compMod_lodes, aes(x=x, stratum=stratum, fill=stratum, alluvium=alluvium, y=freq))+geom_flow()+geom_stratum()+scale_y_continuous(name='Amplicon sequence variants \n (273 total)')+scale_x_discrete(name='', labels=c('Hellinger \n network', 'Center log ratio \n network'))+scale_fill_discrete(name='Network modules', labels=paste0(c(rep('Hellinger & CLR: ', 2), 'Hellinger: ', 'CLR: ', 'Hellinger: ', 'CLR: ', 'CLR: ', 'Hellinger: ', 'CLR: '), levels(compMod_lodes$stratum)))+theme_classic()+theme(legend.text = element_text(face='italic'))

clr_greenyellowbrown<-ggplot(clr_network$module.data%>%filter(Visit %in% c(3, 5, 7, 8, 9, 10) & moduleColor %in% c("green", 'yellow', 'brown')), aes(x=as.factor(Visit), y=modAbund, color=as.factor(Case)))+stat_summary(geom='point', fun.y=mean)+facet_wrap(module_label3~., ncol=1, scales='free_y')+
    stat_summary(fun.data = mean_cl_boot, geom='errorbar', width=0.5)+#stat_compare_means(aes(label = ..p.signif..), size=5, hide.ns = TRUE, label.y=0.50)+
    geom_line(aes(group=UniquePersonID), alpha=0.1)+geom_smooth(aes(group=as.factor(Case), color=as.factor(Case)), se=F)+scale_color_manual(values=c("darkorange3", "royalblue3"), labels=c("ECC case", "Control"), name="")+theme_classic()+
    scale_x_discrete(name="Age (months)", labels=visit_labels2, expand=c(.1, .1))+ylab("Summed module relative abundance")+
    theme(text=element_text(size=12))+theme(axis.text.x=element_text(angle = 45, vjust = 0.5, hjust=.5))+
    #scale_y_continuous(breaks=c(0, .5, 1), minor_breaks = c(.25, .75))+
    theme(legend.position='bottom')+theme(strip.text = element_text(face='italic'))

hellinger_yellowbrowngreen<-ggplot(hellinger_network$module.data%>%filter(Visit %in% c(3, 5, 7, 8, 9, 10) & moduleColor %in% c("green", 'yellow', 'brown')), aes(x=as.factor(Visit), y=modAbund, color=as.factor(Case)))+stat_summary(geom='point', fun.y=mean)+facet_wrap(module_label3~., ncol=1, scales='free_y')+
    stat_summary(fun.data = mean_cl_boot, geom='errorbar', width=0.5)+#stat_compare_means(aes(label = ..p.signif..), size=5, hide.ns = TRUE, label.y=0.50)+
    geom_line(aes(group=UniquePersonID), alpha=0.1)+geom_smooth(aes(group=as.factor(Case), color=as.factor(Case)), se=F)+scale_color_manual(values=c("darkorange3", "royalblue3"), labels=c("ECC case", "Control"), name="")+theme_classic()+
    scale_x_discrete(name="Age (months)", labels=visit_labels2, expand=c(.1, .1))+ylab("Summed module relative abundance")+
    theme(text=element_text(size=12))+theme(axis.text.x=element_text(angle = 45, vjust = 0.5, hjust=.5))+
    #scale_y_continuous(breaks=c(0, .5, 1), minor_breaks = c(.25, .75))+
    theme(legend.position='bottom')+theme(strip.text = element_text(face='italic'))

network_sens2<-plot_grid(hellinger_yellowbrowngreen+stat_compare_means(label='p.signif', hide.ns=T)+ggtitle('Hellinger networks'), clr_greenyellowbrown+stat_compare_means(label='p.signif', hide.ns=T)+ggtitle('CLR networks'))
```

```{r FIG 12, include=F}
network_sensall<-plot_grid(network_sensitivity, network_sens2, labels=c('A', 'B'))
cairo_pdf(file.path(supp.out, 'robust_network.pdf'), family='Arial', width=20, height=8)
network_sensall
dev.off()
```

```{r ami ari ri}
#AMI, ARI, RI between 
clr_hellinger_cohra=clustComp(compareModules$clr, compareModules$hellinger)%>%bind_rows()%>%mutate(comparison='clr vs hellinger', cohort='COHRA2')

compareModules_trim=compareModules%>%filter(
  clr %in% c("Prevotella histicola & Rothia ASV49 network\n (central taxa: Megasphaera micronuciformis)", "Haemophilus parainfluenzae & Neisseria ASV9 network\n (central taxa: Fusobacterium periodonticum)", "Veillonella sp._HMT_780 & Granulicatella elegans network\n (central taxa: Alloprevotella sp._HMT_473)") & 
  hellinger %in% c("Veillonella ASV5 & Streptococcus ASV8 network\n (central taxa: Lachnoanaerobaculum orale)" , "Veillonella sp._HMT_780 & Granulicatella elegans network\n (central taxa: Alloprevotella sp._HMT_473)", "Haemophilus parainfluenzae & Neisseria ASV9 network\n (central taxa: Fusobacterium periodonticum)"))
clr_hellinger_cohra_trim=clustComp(compareModules_trim$clr, compareModules_trim$hellinger)%>%bind_rows()%>%mutate(comparison='clr vs hellinger: trimmed networks', cohort='COHRA2')

comparedf=rbind(comparedf, clr_hellinger_cohra, clr_hellinger_cohra_trim)

compare_cluster_table=comparedf%>%select(cohort, comparison, RI, ARI, AMI)
```

```{r TABLE 7}
write.xlsx(compare_cluster_table, file=file.path(supp.out, 'SupplementalTable7.xlsx'), overwrite = T)
```

## Replication analysis 

Supplemental figure 13 shows the network modules from the Holgerson et al. cohort which were similar in composition to the ECC-associated networks in the COHRA2 cohort. Figure 14 shows other network modules from the Holgerson et al. cohort. 

```{r network and cst replicability}
prj_files<-c(list.files(file.path(data.out, 'PRJ', 'Network'), full.names = T), list.files(file.path(data.out, 'PRJ', 'CST'), full.names = T)[-4])
prj=new.env()
prj_files%>%purrr::map(function(file){load(file, envir = prj)})

load(file.path(data.out, 'PRJ', 'phyloseq', 'phyasv_postTax.Rdata'), envir = prj)
prj$ps_long=microbiome::transform(prj$prj.filter, 'compositional')%>%psmelt()%>%mutate(ScientificName=paste(Genus, ifelse(is.na(Species), OTU, Species), sep=' '))

nw<-prj$hellinger_network
my_df=prj$ps_long%>% group_by(OTU)%>%dplyr::summarise(m=mean(Abundance))%>%mutate(name=OTU)
source(file.path(code.dir, 'Analysis', 'createTidyGraph.R'))


br_yel_cohra<-hellinger_network$intraStats%>%filter(module%in% c('brown', 'yellow'))%>%pull(ScientificName)
by_cohra=br_yel_cohra[!str_detect(br_yel_cohra, 'ASV')]

matching_networks<-gg_whole%>%activate(nodes)%>%filter(module %in% c('turquoise', 'brown')) %>%ggraph('nicely')+geom_edge_link(aes(width=weight), alpha=0.8, color='grey')+geom_node_point(aes(color=GenusLabel, size=m))+geom_node_text(aes(label=ifelse(m>0.01, ScientificNameLabel, ifelse(ScientificName %in% by_cohra, ScientificNameLabel, ''))), fontface='italic', family='Times', repel=TRUE, size=3)+guides(size=FALSE, edge_color=FALSE, color=guide_legend(nrow=2,byrow=TRUE), edge_width=FALSE)+scale_color_manual('', values=genus_colors)+theme_graph(base_family='sans')+theme(legend.position = 'bottom')+scale_edge_width(range = c(0.1, 2))+theme(strip.text.x = element_blank())+facet_nodes(~module, scales='free')

other_networks<-gg_whole%>%activate(nodes)%>%filter(!(module %in% c('turquoise', 'brown'))) %>%ggraph('nicely')+geom_edge_link(aes(width=weight), alpha=0.8, color='grey')+geom_node_point(aes(color=GenusLabel, size=m))+geom_node_text(aes(label=ifelse(m>0.01, ScientificNameLabel, ifelse(ScientificName %in% by_cohra, ScientificNameLabel, ''))), fontface='italic', family='Times', repel=TRUE, size=3)+guides(size=FALSE, edge_color=FALSE, color=guide_legend(nrow=2,byrow=TRUE), edge_width=FALSE)+scale_color_manual('', values=genus_colors)+theme_graph(base_family='sans')+theme(legend.position = 'bottom')+scale_edge_width(range = c(0.1, 2))+theme(strip.text.x = element_blank())+facet_nodes(~module, scales='free')

cohra_names=hellinger_network$intraStats%>%select(module, ScientificName)%>%filter(!str_detect(ScientificName, 'ASV'))%>%group_by(module)%>%nest()

prj_names=prj$hellinger_network$intraStats%>%select(module, ScientificName)%>%filter(!str_detect(ScientificName, 'ASV'))%>%group_by(module)%>%nest()

shared_taxa=data.frame('prj'=character(), 'cohra'=character(), 'common'=character(), 'different'=character(), 
                       n_common=integer(), n_different=integer(), n_cohra=integer(), n_prj=integer())
for( i in 1:length(cohra_names$data)){
  for(j in 1: length(prj_names$data)){
    common=intersect(prj_names$data[[j]]$ScientificName, cohra_names$data[[i]]$ScientificName)
    n_common=length(common)
    different=setdiff(prj_names$data[[j]]$ScientificName, cohra_names$data[[i]]$ScientificName)
    n_diff=length(different)
    n_cohra=nrow(cohra_names$data[[i]]); n_prj=nrow(prj_names$data[[j]])
    new_row=data.frame(prj_names$module[[j]], cohra_names$module[[i]], paste(common, collapse='; '), 
              paste(different, collapse='; '), n_common, n_diff, n_cohra, n_prj)
    names(new_row)=names(shared_taxa)
    shared_taxa=rbind(shared_taxa, new_row)
  }
}
shared_taxa=shared_taxa%>%mutate(p_sharedC=n_common/n_cohra, p_sharedP=n_common/n_prj)

bVSb=shared_taxa %>%filter(cohra=='brown'& prj=='brown')
yVSt=shared_taxa %>%filter(cohra=='yellow'& prj=='turquoise')
```

```{r FIG 13, include=F}
matching_networks
cairo_pdf(file.path(supp.out, 'SupplementalFigure13.pdf'), family='Arial', width=18, height=10)
matching_networks
dev.off()
```

```{r FIG 14, include=F}
other_networks
cairo_pdf(file.path(supp.out, 'SupplementalFigure14.pdf'), family='Arial', width=18, height=10)
other_networks
dev.off()
```

```{r ami ari for between cohorts}
named_asvs=left_join(hellinger_network$intraStats%>%filter(!is.na(Species))%>%mutate(moduleCohra=paste0(module, 'cohra'))%>%select(ScientificName, moduleCohra), prj$hellinger_network$intraStats%>%filter(!is.na(Species))%>%mutate(modulePRJ=paste0(module, '_prj'))%>%select(ScientificName, modulePRJ))

named_asvs_nona=named_asvs%>%filter(!is.na(modulePRJ) & !is.na(moduleCohra))

named_asvs_trim=named_asvs%>%filter(modulePRJ %in%c('blue_prj', 'brown_prj', 'turquoise_prj') & moduleCohra %in% c('', ''))
```

 To compare the relatedness of the amplicon sequence variants assigned to various network modules across the COHRA and Holgerson cohort, we performed multiple sequence alignment of the amplicons using the R packages msa, using the ClustalW algorithim. We computed pairwise distances from the DNA sequences using the r function dist.dml from the r package phangorn, using the JC69 model. We created a neighbor joining tree using the phangorn function NJ, then fit a generalized time-reversible with gamma rate maximum likelihood tree using the neighbor joining tree as a starting point. We obtained 100 bootstrap values for the tree using bootstrap.pml and plotted the tree using ggtree and collapsed branches present in <50 of the bootstrapped trees. Supplemental Figure 15 and 16 show bootstrapped phylogenetic trees for amplicons belonging to the *Neisseria* and *Veillonella* generas (respectively) from the Holgerson *et al.* and COHRA2 cohorts.

```{r network trees create data}
ps.f<-ps_meta_nodups
prj.f<-prj$prj.filter
#rename taxa
taxa_names(ps.f)<-paste0(taxa_names(ps.f), '_COHRA')
taxa_names(prj.f)<-paste0(taxa_names(prj.f), '_PRJ')
cohraasv<-taxa_names(ps.f)
prjasv<-taxa_names(prj.f)
#make new taxa frame
tax_table_merged<-as.matrix(rbind(left_join(as.data.frame(cbind(
  ps.f@tax_table, 'moduleColor'=hellinger_network$taxaColors, 'source'='COHRA')), hellinger_network$networkNames %>% select(moduleColor, module_label3))%>%
    mutate(modulesource=paste0(source,': ', module_label3)), left_join(as.data.frame(cbind(
      prj.f@tax_table, 'moduleColor'=prj$hellinger_network$taxaColors, 'source'="Holgerson et al.")), prj$hellinger_network$networkNames %>%
        select(moduleColor,module_label3))%>%
    mutate(modulesource=paste0(source,': ', module_label3))))

rownames(tax_table_merged)<-c(cohraasv, prjasv)
refseq_merged<-c(ps.f@refseq, prj.f@refseq)
otu_merged<-data.frame(matrix(rep(1, 701*4), nrow=4, ncol=701), row.names=paste0(rep('FAKE_', 4), 1:4))
sample_merged<-data.frame(matrix(rep(0, 4*4), nrow=4, ncol=4), row.names=paste0(rep('FAKE_', 4), 1:4))
names(otu_merged)<-c(taxa_names(ps.f), taxa_names(prj.f))
ps_relatedness<-phyloseq(otu_table(otu_merged, taxa_are_rows=F), tax_table(tax_table_merged), sample_data(sample_merged), refseq(refseq_merged))
```

```{r tree function and set seed, eval=F}
set.seed(03172021)
#this takes a long time to run, so ran once and saved plot as robject to laod back in. 
#make sure to run again if substantial changes up stream
#function based off of code here: 
#https://bioconductor.org/help/course-materials/2017/BioC2017/Day1/Workshops/Microbiome/MicrobiomeWorkflowII.html#supervised_learning
#https://fuzzyatelin.github.io/bioanth-stats/module-24/module-24.html
#fn function from https://stackoverflow.com/questions/67705772/non-standard-evaluation-does-not-work-error-in-hsimpleerrormsg-call
fn <- function(ps, level, choice) {
        x <- paste0(level,"==","'",choice,"'")
        oldTax <- data.frame(tax_table(ps))
        newTax <- subset(oldTax, eval(parse(text=x)))
        tax_table(ps) <- tax_table(as.matrix(newTax))
        return(ps)
}

make_myTree=function(ps_object, level, choice, prob){
  g=fn(ps_object, level, choice)
  seqs=g@refseq
  #multiple sequence alignment
  mult=msa(seqs, method="ClustalW", type="dna", order="input")
  mult_dnabin<-as.DNAbin(mult)
  phang.align <- as.phyDat(mult, type="DNA", names=getSequence(seqs))
  dm <- dist.ml(phang.align) #distance ml for DNA
  dd<-dist.dna(mult_dnabin, model='N', pairwise.deletion = F)
  treeNJ_dd<-NJ(dd)
  treeNJ <- NJ(dm) # Note, tip order != sequence order
  fit = pml(treeNJ, data=phang.align)
  fitGTR <- update(fit, k=4, inv=0.2)
  fitGTR <- optim.pml(fitGTR, model="GTR", optInv=TRUE, optGamma=TRUE, rearrangement = "stochastic", control = pml.control(trace = 0))
  #bootstrap tree
  bs=bootstrap.pml(fitGTR, bs=100, optNni=TRUE, control=pml.control(trace=0), multicore = T)
  s<-plotBS(midpoint(fitGTR$tree), bs, p=prob, type='p')
  s2=ggtree::as.polytomy(s, feature='node.label', fun=function(x) as.numeric(x)<prob)
  #add tree to phyloseq object
  tree_collapsed<-merge_phyloseq(g, s2)
  return(tree_collapsed)
}


neisseria_tree=make_myTree(ps_relatedness, 'Genus',  'Neisseria', prob=50)

neis_tree<-ggtree(neisseria_tree)+geom_tiplab(aes(label=ifelse(label%in% c('ASV9_COHRA', 'ASV12_COHRA'), label, Species), color=modulesource))+geom_point(aes(color=modulesource), size=2)+scale_color_manual('Study source & network assignment', values=c("cornflowerblue", 'darkorange', "grey", "grey20", "blue4"))+theme(legend.position = c(0.7, 0.2))+theme_tree2()
d<-neis_tree$data
d<-d[!d$isTip,]
neis_tree_collapsed<-neis_tree+geom_text(data=d, aes(label=label), hjust=1.25)+theme(text=element_text(face='italic'))

veillonella_tree=make_myTree(ps_relatedness, 'Genus', 'Veillonella', prob=50)

veil_tree<-ggtree(veillonella_tree)+geom_tiplab(aes(label=ifelse(label %in% c('ASV4_COHRA', 'ASV5_COHRA'), label, Species), color=modulesource))+geom_point(aes(color=modulesource), size=2)+scale_color_manual('Study source & network assignment', values=c("grey", "darkgoldenrod", 'green', 'darkgreen', 'grey5', "grey20", 'darkorange'))+theme(legend.position = c(0.7, 0.2))+theme_tree2()
d<-veil_tree$data
d<-d[!d$isTip,]
veil_tree_collapsed<-veil_tree+geom_text(data=d, aes(label=label), hjust=1.25)+theme(text=element_text(face='italic'))

tree_figure=plot_grid(neis_tree_collapsed, veil_tree_collapsed, ncol=1, labels=c('A', 'B'))

save(tree_figure, file=file.path(og.data, 'treefigure.Rdata'))
```

```{r FIG 15, include=F}
load(file=file.path(og.data, 'treefigure.Rdata'))
tree_figure
```

```{r FIG 15 save}
ggsave(tree_figure, file=file.path(supp.out, 'SupplementalFigure15.pdf'), height=12, width=25)
```


#ECC-associated guilds associate with concurrent salivary pH and future *S. mutans* prevalence 

Salivary pH alone did not distinguish cases and controls at the 12-month and 24-month visits (Supplemental Figure 16A). However, the *H. parainfluenzae-Neisseria* ASV9 network module was positively correlated with salivary pH at the 12- and 24-month visits (Supplemental Figure 16B) and a similar trend existed between salivary pH and the community state types (Supplemental Figure 16D). 

```{r salivary pH future smutans}
#create cst group variable for plotting
netcst=left_join(left_join(hellinger_network$module.data, 
                               left_join(main_cst, cst_recodes)), MV)%>%
                  left_join(div%>%
                            dplyr::select(FoxCavID, Chao1, Shannon, smutAbundance,
                            smutPrevalence))%>%
        mutate(CSTGroup=case_when(CST.8%in% c(1, 2)~'Late childhood CSTs', 
                                  CST.8 %in% c(3, 4)~"Infant CSTs", 
                                  CST.8 %in% c(5, 6)~'Early childhood CSTs'),
               )
netcst.57=netcst%>%filter(Visit %in% c(5, 7))%>%mutate(SalivapH=ifelse(SalivapH<2.5, NA, SalivapH))

#salivary pH, concurrent
#module -> ph
net_ph<-ggplot(netcst.57, aes(x=SalivapH, y=modAbund))+geom_point()+facet_grid(Visit~module_label3, labeller=labeller(Visit=visit_labels2))+geom_smooth(color='red', method='lm')+theme_classic()+stat_cor()+ggtitle('Salivary pH & WCN module abundance')+theme(strip.text.x = element_text(face='italic'))+xlab('Salivary pH')+ylab('Summed module relative abundance')+theme(text=element_text(size=14))

cstph=ggplot(netcst.57%>%filter(moduleColor=='brown' & !is.na(CSTlabels)), aes(x=CSTlabels, y=SalivapH, color=CSTlabels))+geom_boxplot()+geom_jitter(alpha=0.5)+facet_grid(Visit~., labeller=labeller(Visit=visit_labels2))+stat_compare_means(label.y=8.5)+theme_classic()+theme(axis.text.x = element_blank(), axis.title.x = element_blank())+theme(legend.position = 'bottom', legend.text = element_text(face='italic'))+scale_color_manual(name="", values=c("thistle","lightcoral", "deepskyblue",  "darkorange", "royalblue2", "darkorange3",  "royalblue4", "darkorange4", 'grey'), guide=guide_legend(nrow=2))+theme(legend.position = 'bottom')+stat_compare_means(label.y=8.5, label='p.format', comparisons=list(c("Gemella ASV2 - H. parainfluenzae - Neisseria ASV9", "Streptococcus ASV8 - Neisseria ASV12"), c("H. parainfluenzae - Neisseria ASV9", "Neisseria ASV12 - V. dispar"), c("Streptococcus ASV1 dominated w/ Gemella ASV2", "Streptoccous ASV1 dominated w/ G. elegans" )))+ggtitle('Salivary pH & CST')+theme(text=element_text(size=14))

#lagged data for s mutans 
netcst.lag<-netcst %>%group_by(BabySubjectID, moduleColor)%>%arrange(BabySubjectID, AgeAtExamMonths, moduleColor)%>%mutate(modAbund.lag=lag(modAbund, n=1, default=NA, order_by=BabySubjectID), SalivapH.lag=lag(SalivapH, n=1, default=NA, order_by=BabySubjectID), smut.lag=lag(smutPrevalence, n=1, default=NA, order_by = BabySubjectID), CST.lag=lag(CSTlabels, n=1, default=NA, order_by=BabySubjectID), 
Dentobuff.lag=lag(Dentobuff, n=1, order_by=BabySubjectID, default = NA))

smut_lag<-ggplot(netcst.lag %>% filter(moduleColor%in%c('brown', 'yellow')& Visit %in% c(7, 8)), aes(x=smutPrevalence, y=modAbund.lag))+facet_grid(Visit~module_label2, labeller=labeller(Visit=visit_labels2))+geom_boxplot()+theme_classic()+stat_compare_means()+ylab("Summed module abundance in the previous visit")+xlab("S. mutans 16S amplicon detected in saliva")+theme(axis.title.x = element_text(face='italic'))+theme(strip.text.x=element_text(face='italic'))+theme(text = element_text(size=14))

case_ph<-ggplot(MV %>% filter(Visit %in% c(5, 7)), aes(x=Case, y=SalivapH))+geom_boxplot()+stat_compare_means()+geom_jitter(alpha=0.5)+facet_grid(Visit~., labeller=labeller(Visit=visit_labels2))+theme_classic()+ylab("Salivary pH")+xlab("Early childhood caries case status")+theme(text=element_text(size=14))+theme(legend.position = c(0.1, 0.1))

ph_smut<-plot_grid(plot_grid(case_ph, net_ph, ncol=2, rel_widths=c(0.2, 0.8), labels=c('A', 'B')), plot_grid(smut_lag, cstph, ncol=2, rel_widths=c(0.35, 0.65), labels=c('C', 'D')), ncol=1)
```

```{r FIG 16, include=F}
ggsave(ph_smut, file=file.path(supp.out, 'SupplementalFigure16.pdf'), height=12, width=25)
```

```{r pull values}
#cor function & purr mapping for correlations between network module abundance and pH  
cor_fun<-function(df, var1, var2){
  cor.test(df[[var1]], df[[var2]], method='pearson')%>%tidy()
}
hellinger_nest<-group_by(netcst.57, Visit, module_label3)%>%nest()
hellinger_nest<- hellinger_nest %>% mutate(Saliva_pH=purrr::map(data, cor_fun, var1='SalivapH', var2='modAbund'), Primary_Teeth=purrr::map(data, cor_fun, var1='Prim_Tot_Teeth_Present', var2='modAbund'))%>% unnest(cols=c(Saliva_pH, Primary_Teeth), names_sep='.')

epiround<-function(x, n){
  base=1*10^(-n)
  base_char=paste0('<', base)
  ifelse(x<base, base_char, round(x, n))
}
pullRho<-function(name, var, time){
  est=paste0(var, '.estimate')
  pval=paste0(var, '.p.value')
  rho=hellinger_nest %>% filter(str_detect(module_label3, name) & Visit==time)%>%pull(!!est)%>%round(2)
  p=hellinger_nest %>% filter(str_detect(module_label3, name) & Visit==time)%>%pull(!!pval)%>%epiround(3)
  return(paste0('Rho=', rho, '; p-val=', p))
}

#df for pull percent of cst --> smut prevalence
smut_lag_cst<-netcst.lag %>% filter(moduleColor=='brown')%>%group_by(Visit, CST.lag)%>%mutate(n_total=n())%>%group_by(Visit, CST.lag, smutPrevalence)%>%mutate(n=n())%>%mutate(percent=round((n/n_total)*100, 1))%>%select(n, n_total, percent)%>%unique()
```

The groups of co-occurring taxa identified from the unsupervised clustering methods associated with concurrent salivary pH at the 12- and 24-month visits. Increasing abundance of the *H. parainfluenzae & Neisseria ASV9* module was correlated with increasing salivary pH (`r pullRho('Haemo', 'Saliva_pH', 5)`, `r pullRho('Haemo', 'Saliva_pH', 7)` respectively). A similar trend was apparent when comparing mean salivary pH between samples assigned to the different community state types (Supplemental Figure XX). 

Children who went on to have *S. mutans* amplicon detected in their saliva at the next visit had lower abundances of the *H. parainfluenzae & Neisseria ASV9* network and higher abundances of the *Veillonella ASV5 & Streptococcus ASV8 network* network than children who did not go on to have *S. mutans detected* (Supplementary Figure XX). A similar trend existed for community state types. Of the children assigned to the *Gemella ASV2 - H. parainfluenzae - Neisseria ASV9* at the 12-month visit, `r smut_lag_cst %>% filter(Visit==7 & CST.lag=='Gemella ASV2 - H. parainfluenzae - Neisseria ASV9' & smutPrevalence=='Yes')%>%pull(percent)`% of children had detectable *S. mutans* amplicon at the 24-month visit, as compared to `r smut_lag_cst %>% filter(Visit==7 & CST.lag=='Streptococcus ASV8 - Neisseria ASV12' & smutPrevalence=='Yes')%>%pull(percent)`% of the children assigned to the children assigned to the *Streptococcus ASV8 - Neisseria ASV12* CST. Of the children assigned to the *H. parainfluenzae - Neisseria ASV9* CST at the 24-month visit, `r smut_lag_cst %>% filter(Visit==8 & CST.lag=='H. parainfluenzae - Neisseria ASV9' & smutPrevalence=='Yes')%>%pull(percent)`% of children had detectable *S. mutans* amplicon at the 36-month visit, as compared to `r smut_lag_cst %>% filter(Visit==8 & CST.lag=='Neisseria ASV12 - V. dispar' & smutPrevalence=='Yes')%>%pull(percent)`% of children assigned to the *Neisseria ASV12 - Veillonella ASV5*.


The *H. parainfluenzae & Neisseria ASV9* module did not correlate with the total number of primary teeth present at the 12- or 24-month visits, although the *Streptococcus ASV1 and Neisseria ASV12 module* did (Supplemental Figure 17). Similarly, although the mean number of primary teeth increased between community state types which were dominate at different times, the number of primary teeth did not significantly differ between the community state types that were codominant at any one time (Supplemental Figure XX). Among the subset of children recruited from Pennsylvania, data on the approximate age at first tooth emergence was available. In this subset, age at first tooth emergence was not associated with either WCN modules or CSTs.

```{r tooth number tooth emergence}
#cst to primary tooth count
cst_prim<-ggplot(netcst.57%>%filter(moduleColor=='brown' &!is.na(CSTlabels)), aes(x=CSTlabels, y=Prim_Tot_Teeth_Present, color=CSTlabels))+geom_boxplot()+geom_jitter()+facet_grid(Visit~., labeller=labeller(Visit=visit_labels2))+stat_compare_means(label.y=25)+theme_classic()+theme(axis.text.x = element_blank(), axis.title.x = element_blank())+theme(legend.position = 'bottom', legend.text = element_text(face='italic'))+scale_color_manual(name="Community state type", values=c("thistle","lightcoral", "deepskyblue",  "darkorange", "royalblue2", "darkorange3",  "royalblue4", "darkorange4", 'grey'), guide=guide_legend(nrow=2))+stat_compare_means(label.y=22, label='p.signif', comparisons=list(c("Gemella ASV2 - H. parainfluenzae - Neisseria ASV9", "Streptococcus ASV8 - Neisseria ASV12"), c("H. parainfluenzae - Neisseria ASV9", "Neisseria ASV12 - V. dispar"), c("Streptococcus ASV1 dominated w/ Gemella ASV2", "Streptoccous ASV1 dominated w/ G. elegans" )))+ggtitle('Primary teeth present & CST')+ylab('Primary teeth present (count)')+theme(text=element_text(size=14))
#module -> teeth present
net_teeth<-ggplot(netcst.57, aes(x=Prim_Tot_Teeth_Present, y=modAbund))+geom_point()+facet_grid(Visit~module_label3, labeller = labeller(Visit=visit_labels2))+geom_smooth(color='red', method='lm')+theme_classic()+stat_cor()+ggtitle('Primary teeth present & WCN module abundance')+theme(strip.text.x = element_text(face='italic'))+xlab('# primary teeth')+ylab('Summed module relative abundance')+theme(text=element_text(size=14))
#case to teeth
case_teeth<-ggplot(MV%>%filter(Visit %in% c(5, 7)), aes(x=Case, y=Prim_Tot_Teeth_Present, color=Case))+geom_violin()+geom_boxplot(width=0.2)+geom_jitter()+facet_grid(Visit~., labeller=labeller(Visit=visit_labels2))+theme_classic()+stat_compare_means(label.y=25)+ggtitle('Primary teeth present & case status')+ylab('Primary teeth present (count)')+theme(text=element_text(size=14))+theme(legend.position='none')

prim_teeth<-plot_grid(plot_grid(case_teeth, cst_prim, nrow=1, labels=c('A', 'B'), rel_widths=c(0.3, 0.7), align='h', axis='tb'), net_teeth, nrow=2, labels=c('', 'C'))
```

```{r FIG 17, include=F}
ggsave(prim_teeth, file=file.path(supp.out, 'SupplementalFigure17.pdf'), height=12, width=25)
prim_teeth
```

# Whole-genome shotgun metagenomics of incident samples reveals taxonomic and functional differences between incident case- and control-samples

## Volcano plot: taxa with abundances different between cases and controls in whole genome metagenome samples 

```{r volcano plot taxa}
#load data
load(file.path(data.out, 'MS', 'deseqresults.Rdata'))

cbPalette_volc=c('red', "#E69F00", "#56B4E9", "#009E73", "darkblue", "#0072B2", "purple", "#CC79A7", "#00FF00", 'darkred', "#000000")
taxa_of_interest=c('Neisseria perflava', 'Neisseria mucosa', 'Haemophilus parainfluenzae', 'Fusobacterium periodonticum', 'Fusobacterium nucleatum', 'Streptococcus sangunis', 'Prevotella salivae', 'Actinomyces graevenitzii', 'Streptococcus vestibularis', 'Prevotella histicola')
t3=t3%>%mutate(label=ifelse(Taxa %in% taxa_of_interest & padj<0.05, Taxa, ifelse((-log10(pvalue)>10)|(-log10(pvalue)>5 & SampleType=='saliva'), Taxa, '')))
taxa_volc=ggplot(t3%>%filter(Taxa!='Unmapped'),
       aes(x=log2FoldChange, y=-log10(pvalue),
           label=label,color=ColorCode))+
    geom_point()+geom_text_repel(max.overlaps = 500)+
    facet_wrap(~SampleType, scales='free_x')+
    scale_color_manual('', values=cbPalette_volc, na.translate=FALSE)+
    theme_classic()+xlab('\n \n log2 Fold change ')+
    theme(axis.line.x = element_line(arrow = 
                                         grid::arrow(length = unit(0.3, "cm"), ends = "both")))+
    theme(legend.position='bottom')

ggsave(taxa_volc, filename=file.path(supp.out, 'taxaVolcano.pdf'), units=c('in'), width=18, height=8)

t3_table=t3%>%select(-ColorCode)%>%arrange(padj)

```

```{r TABLE 8}
write.xlsx(t3_table, file=file.path(supp.out, 'SupplementalTable8.xlsx'), overwrite=T)
```

```{r FIG 18, include=F}
taxa_volc
ggsave(taxa_volc, filename=file.path(supp.out, 'SupplementalFigure18.pdf'), units=c('in'), width=18, height=8)
```

```{r Table 9}
r3_table=r3%>%select(SampleType, KEGG, kname, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj, PathAnnotation)%>%arrange(padj)
write.xlsx(r3_table, file=file.path(supp.out, 'FunctionsDeseq2.xlsx'), overwrite = T)
```

Significant differences also existed in the abundance of KEGG pathways in both plaque and saliva samples by case status (Figure 3A; Supplemental Table 9). Among the top 20 most significantly different KEGG orthologs were additional KEGG orthologs with annotations to antibiotic production, antibiotic resistance and quorum sensing. Many KEGG orthologs significantly enriched in plaque samples annotated to metabolic or environmental information processing, including ABC transporters, transcription factors, and two component systems (Figure 3B). 

Functional differences sometimes reflect taxonomic differences. For example, all the case enriched KEGG orthologs annotating to oxidative phosphorylation were found only in the genus Candida (Supplemental Table 10 and Supplemental Figure 19). Candida was only present in case plaque samples. 



```{r oxidative phosphorylation exploration}
load(file.path(data.out, 'MS', 'taxpathannotation.Rdata'))

op_case=bind_rows(op_taxa_list, .id='KEGG')%>%left_join(species_info$sample_table)%>%mutate(caseass=ifelse(KEGG %in% 'V/A-type', 'Control associated', 'Case associated'))%>%group_by(KEGG, SampleName)%>%filter(caseass=='Case associated')%>%ggplot(aes(x=Species, y=Abundance, color=CaseStatus))+geom_boxplot()+facet_wrap(~KEGG)+coord_flip()+theme_classic()+theme(axis.text.y = element_text(face='italic'))+scale_color_manual(values=c("darkorange3", "royalblue3"), labels=c("ECC case", "Control"), name="")

op_control=bind_rows(op_taxa_list, .id='KEGG')%>%left_join(species_info$sample_table)%>%mutate(caseass=ifelse(KEGG %in% 'V/A-type', 'Control associated', 'Case associated'))%>%group_by(KEGG, SampleName)%>%filter(caseass=='Control associated')%>%ggplot(aes(x=Species, y=Abundance, color=CaseStatus))+geom_boxplot()+facet_wrap(~KEGG)+coord_flip()+theme_classic()+theme(axis.text.y = element_text(face='italic'))+scale_color_manual(values=c("darkorange3", "royalblue3"), labels=c("ECC case", "Control"), name="")

op_plot=plot_grid(op_case, op_control, labels=c('A', 'B'), ncol=1)

op_table=left_join(r3%>%filter(padj<0.05 & PathAnnotation_01_3=='Oxidative phosphorylation')%>%mutate(p=word(kname, 1, 3))%>%mutate(p=word(p, 1))%>%select(KEGG, kname, p, log2FoldChange), lapply(op_taxa_list, function(x) x %>%select(Species)%>%unique())%>%bind_rows(.id='p')%>%group_by(p)%>%dplyr::summarise(Taxa=list(Species)))

```

```{r FIG 19, include=F}
ggsave(op_plot, filename=file.path(supp.out, 'oSupplementalFigure19.pdf'), units=c('in'), width=10, height=8)
op_plot
```

```{r Table 10}
write.xlsx(op_table, file=file.path(supp.out, 'SupplementalTable9.xlsx'), overwrite = T)
```


